Spatiotemporal model of `r paste(params$species)` occupancy

### 2020-11-26 analysis/sand-lance/biomass-density-model.Rmd
working occupancy model

---
title: "Spatiotemporal model of `r paste(params$species)` biomass"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    covs: "-sand-depth-temp"
    # covs: "-mud-depth-temp"
    # covs: "-mud-sand-depth-temp"
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, by = shallow, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}
if(params$covs == "-mud-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

```

@@

Check for substrate correlations -- sandy somewhat correlated with depth
```{r eval= FALSE}
ggplot(data, aes(raw_depth, sandy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

ggplot(data, aes(raw_depth, sandy-muddy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

@@

ggplot(filter(data, raw_depth < 100), aes(depth, sandy, colour = depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1) 
```

@@

ggsave(paste0("figs/depth-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_depth})
```

@@

try({g_roms})
```

@@

ggsave(paste0("figs/temp-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_temp})
```

@@

ggsave(paste0("figs/sand-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_sand})
```

@@

ggsave(paste0("figs/mud-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_mud})
```

@@

  nd_all <- nd_all %>% mutate(
    depth_scaled = (log(depth) - depth_mean)/depth_sd,
    sandy_scaled = (sandy - sandy_mean)/sandy_sd,
    muddy_scaled = (muddy - muddy_mean)/muddy_sd,
    roms_scaled = (roms - roms_mean)/roms_sd,
    temp_scaled = (temp - temp_mean)/temp_sd
  ) 
  nd_all <- nd_all %>% filter(year > 2003) %>% 
  # merge survey pairs such that even year surveys are pooled w the previous odd year
    mutate(year_pair = (2 * round((year+0.8)/2))-1) %>% 
    # 0.8 is required to avoid 5s rounding to even in R
    filter(year < 2019)

@@

  try({
    predicted <- readRDS(paste0("data/", spp, "/predictions-", spp, 
      covs, "-", ssid_string, "-tot-biomass-ar1-", 
      params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))
  })

>>>>

---
title: "Spatiotemporal model of `r paste(params$species)` biomass"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    # covs: "-mud-depth-temp"
    # covs: "-mud-sand-depth-temp"
    covs: "-sand-x-depth-temp" # doesn't converge
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(sandy_scaled, by = shallow, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 2) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  muddy_scaled + 
  sandy_scaled + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-x-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  sandy_scaled + 
  s(depth_scaled, k = 3) + 
  te(depth_scaled, sandy_scaled, k = 3, m=1) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}
```

@@

Check for substrate correlations -- sandy somewhat correlated with depth
```{r}
ggplot(data, aes(raw_depth, sandy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

ggplot(filter(data, raw_depth < 100), aes(raw_depth, sandy, colour = raw_depth)) + 
  geom_point(alpha=0.75, size=0.5) +
  scale_color_viridis_c(trans=fourth_root_power, direction = -1)

@@

# ggplot(data, aes(raw_depth, sandy-muddy, colour = raw_depth)) + 
#   geom_point(alpha=0.75, size=0.5) +
#   scale_color_viridis_c(trans=fourth_root_power, direction = -1) 
```

@@

ggsave(paste0("figs/depth-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_depth}, silent = TRUE)
```

@@

try({g_roms}, silent = TRUE)
```

@@

ggsave(paste0("figs/temp-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_temp}, silent = TRUE)
```

@@

ggsave(paste0("figs/sand-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_sand}, silent = TRUE)
```

@@

ggsave(paste0("figs/mud-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_mud}, silent = TRUE)
```

@@

  nd_all <- nd_all %>% mutate(
    depth_scaled = (log(depth) - depth_mean)/depth_sd,
    sandy_scaled = (sandy - sandy_mean)/sandy_sd,
    muddy_scaled = (muddy - muddy_mean)/muddy_sd,
    roms_scaled = (roms - roms_mean)/roms_sd,
    temp_scaled = (temp - temp_mean)/temp_sd
  ) 
  nd_all <- nd_all %>% filter(year > 2003) %>% 
  # merge survey pairs such that even year surveys are pooled w the previous odd year
    # 0.8 is required to avoid 5s rounding to even in R
    mutate(year_pair = (2 * round((year+0.8)/2))-1,
      shallow = if_else(depth < 100, 1, 0)) %>% 
    filter(year < 2019)

@@

  predicted <- readRDS(paste0("data/", spp, "/predictions-", spp, 
      covs, "-", ssid_string, "-tot-biomass-ar1-", 
      params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))

### 2020-11-26 analysis/sand-lance/occupancy-model.Rmd
working occupancy model

---
title: "Spatiotemporal model of `r paste(params$species)` occupancy"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    covs: "-mud-depth-temp"
    # covs: "-sand-mud-depth-temp"
    # covs: "-sand-by-depth-mud-temp"
    REML: FALSE
    AR1: FALSE # true doesn't seem to work with occupancy models
    fixed_spatial: FALSE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

```{r formula}
if(params$covs == "-sand-depth-roms"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

@@

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-by-depth-mud-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  te(sandy_scaled, depth_scaled, k = 3) + 
  s(muddy_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

```

@@

d <- left_join(data, bath) %>% mutate(
  depth_sam = depth,# rename sensor depth
  depth = coalesce(depth, depth_bath),# use bath depth if sensor depth missing
  raw_depth = depth, # create 'depth' var that's actually log depth for scaling
  log_depth = log(depth), depth = log(raw_depth)) %>% 
  rename(temp = temperature_c, do = do_mlpl)

@@

```{r}
g2 <- ggplot(pred, 
      aes(est, residuals, colour = as.factor(present))) +
    geom_point(alpha = 0.2) + # scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g2
```

@@

nd_depth <- data.frame(
  depth_scaled = seq(min(data$depth_scaled), 
    max(data$depth_scaled), length.out = 30), 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = 0,
  roms_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

ggsave(paste0("figs/depth-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_depth})
```

```{r}
if(clim_cov == "roms"){
nd_roms <- data.frame(
  depth_scaled = min(data$depth_scaled), # use min depth for estimated prob at most relevant depth
  sandy_scaled = 0,
  muddy_scaled = 0,
  roms_scaled = seq(min(data$roms_scaled), 
    max(data$roms_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_roms <- predict(m, newdata = nd_roms, se_fit = TRUE, re_form = NA)
p_roms$roms <- (p_roms$roms_scaled*data$roms_sd[1])+ data$roms_mean[1]

@@

ggsave(paste0("figs/roms-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_roms})
```

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

@@

ggsave(paste0("figs/temp-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_temp})
```

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

ggsave(paste0("figs/sand-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_sand})
```

```{r}
if(covs %in% c("-mud-depth-roms", "-mud-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_mud <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  muddy_scaled = seq(min(data$muddy_scaled), 
    max(data$muddy_scaled), length.out = 50),
  sandy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

ggsave(paste0("figs/mud-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_mud})
```

@@

  try({
    predicted <- readRDS(paste0("data/", spp, "/predictions-", spp, 
      covs, "-", ssid_string, "-tot-binomial-ar1-", 
      params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))
  })

@@

p_mean <- predictions  %>% group_by(X, Y) %>%
  mutate(mean_est = mean(logit2prob(est))) %>% 
  filter(year_pair == 2011) %>% 
  mutate(year = "all years") %>%
  plot_facet_map("mean_est",
    raster_limits = c(0, max_raster),
    transform_col = fourth_root_power,
    legend_position = legend_coords
  ) + labs(fill = "prob") +
  ggtitle(paste0("Estimated occupancy of ", species, 
    " \n(max = ", max_prob, ")"))
print(p_mean)
ggsave(paste0("figs/map-", spp, covs, "-mean-occupancy.png"), width = 5, height = 5)
```

>>>>

---
title: "Spatiotemporal model of `r paste(params$species)` occupancy"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    covs: "-sand-depth-temp"
    # covs: "-mud-depth-temp"
    # covs: "-sand-mud-depth-temp"
    # covs: "-sand-by-depth-mud-temp"
    REML: FALSE
    AR1: FALSE # true doesn't seem to work with occupancy models
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

```{r formula}
if(params$covs == "-sand-depth-roms"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  sandy_scaled + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

@@

if(params$covs == "-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  sandy_scaled + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  muddy_scaled + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  sandy_scaled +
  muddy_scaled + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

# probably too complex to fit
if(params$covs == "-sand-by-depth-mud-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  te(sandy_scaled, depth_scaled, k = 3) + 
  s(muddy_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}
```

@@

d <- left_join(data, bath) %>% mutate(
  depth_sam = depth,# rename sensor depth
  depth = coalesce(depth, depth_bath),# use bath depth if sensor depth missing
  raw_depth = depth, # create 'depth' var that's actually log depth for scaling
  log_depth = log(depth), depth = log(raw_depth),
  shallow = if_else(raw_depth < 100, 1, 0)) %>% 
  rename(temp = temperature_c, do = do_mlpl)

@@

```{r}
g2 <- ggplot(pred, 
      aes(logit2prob(est), residuals, colour = as.factor(present))) +
    geom_point(alpha = 0.2) + # scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g2
```

@@

nd_depth <- data.frame(
  depth_scaled = seq(min(data$depth_scaled), 
    max(data$depth_scaled), length.out = 30), 
  shallow = 1, 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = 0,
  roms_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

ggsave(paste0("figs/depth-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_depth}, silent = TRUE)
```

```{r}
if(clim_cov == "roms"){
nd_roms <- data.frame(
  depth_scaled = min(data$depth_scaled), # use min depth for estimated prob at most relevant depth
  shallow = 1,
  sandy_scaled = 0, 
  muddy_scaled = 0,
  roms_scaled = seq(min(data$roms_scaled), 
    max(data$roms_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_roms <- predict(m, newdata = nd_roms, se_fit = TRUE, re_form = NA)
p_roms$roms <- (p_roms$roms_scaled*data$roms_sd[1])+ data$roms_mean[1]

@@

ggsave(paste0("figs/roms-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_roms}, silent = TRUE)
```

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  shallow = 1, 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

@@

ggsave(paste0("figs/temp-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_temp}, silent = TRUE)
```

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  shallow = 1, 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

ggsave(paste0("figs/sand-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_sand}, silent = TRUE)
```

```{r}
if(covs %in% c("-mud-depth-roms", "-mud-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_mud <- data.frame(
  depth_scaled = min(data$depth_scaled), 
  shallow = 1, 
  muddy_scaled = seq(min(data$muddy_scaled), 
    max(data$muddy_scaled), length.out = 50),
  sandy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

ggsave(paste0("figs/mud-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_mud}, silent = TRUE)
```

@@

predicted <- readRDS(paste0("data/", spp, "/predictions-", spp, 
      covs, "-", ssid_string, "-tot-binomial-ar1-", 
      params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))

@@

p_mean <- predictions  %>% group_by(X, Y) %>%
  mutate(mean_est = mean(logit2prob(est))) %>% 
  filter(year_pair == 2011) %>% 
  mutate(year = "all years") %>%
  plot_facet_map("mean_est",
    raster_limits = c(0, max_raster),
    transform_col = fourth_root_power,
    legend_position = legend_coords
  ) + labs(fill = "prob") +
  ggtitle(paste0("Probability of occupancy by ", species))
print(p_mean)
ggsave(paste0("figs/map-", spp, covs, "-mean-occupancy.png"), width = 5, height = 5)
```
